Modelling the emergence of whisker barrels

### 2019-12-06 paper/briefpaper_stuedits/brief.tex

\begin{abstract}
Brain development relies on an interplay of genetic specification and self-organisation. Striking examples of this interplay can be found in the rodent somatosensory brainstem, thalamus, and cortex, where the arrangement of the facial whiskers is preserved in the arrangement of cellular aggregates to form precise somatotopic maps. We show in simulation how realistic whisker maps can self-organize, by assuming that information is exchanged between adjacent cells only, under the guidance of gene expression gradients. The resulting model provides a simple account of how genetic information can constrain spontaneous pattern formation to faithfully reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$, and axon branches $a(x,t)$, interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^l.
\end{equation}
%
When the final term evaluates to zero, i.e., when the total density of connections sums to one, the density of thalamocortical connections decays at rate $\alpha$. Otherwise the synaptic connection density increases non-linearly ($l>1$) with the axon branching density. Changes in axon branching are modelled as 
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux of axonal branching'. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

The term in parentheses in Eq.~\ref{eq:dc} represents competition between thalamic projections for a limited availability of cortical connections. To introduce competition also in terms of axon branching we introduced
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branching of other projections is high. Divisive normalization is used to keep the axon branch density bounded on each iteration to the initial density  at $t=0$,

>>>>

\begin{abstract}
Brain development relies on an interplay of genetic specification and self-organisation. Striking examples of this interplay can be found in the rodent somatosensory brainstem, thalamus, and cortex, where the arrangement of the facial whiskers is preserved in the arrangement of cell aggregates to form precise somatotopic maps. We show in simulation how realistic whisker maps can self-organize, by assuming that information is exchanged between adjacent cells only, under the guidance of gene expression gradients. The resulting model provides a simple account of how genetic information can constrain spontaneous pattern formation to faithfully reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$, and axon branches $a(x,t)$, interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^l.
\end{equation}
%
Accordingly, when the total density of connections sums to one, the density of thalamocortical connections decays at rate $\alpha$. Otherwise the synaptic connection density increases non-linearly ($l>1$) with the axon branching density. Changes in axon branching are modelled as 
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux of axonal branching'. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

The term in parentheses in Eq.~\ref{eq:dc} represents competition between thalamic projections for a limited availability of cortical connections. To introduce competition also in terms of axon branching we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branching of other projections is high. Divisive normalization is used to keep the axon branch density bounded on each iteration to the initial density  at $t=0$,
